---
title: "AMSS_2025"
author: "Maris Goodwin"
date: "2024-12-29"
output: html_document
---
This is the analysis for my AMMS 2025 presentation.

The abstract is here: 

Title: Understanding seasonality of fish biodiversity using eDNA metabarcoding in Kachemak
Bay, Alaska
Authors: Maris R. Goodwin, Zack Gold, Jennifer L. Tusten, Jessica R. Glass
As global climate change rapidly impacts Arctic and subarctic marine ecosystems, scientists need tools that will rigorously quantify changes in biological communities. Arctic biodiversity monitoring is essential as glacial estuaries face recession, where the dynamic convergence of freshwater and marine environments has historically supported diverse ecological communities. Genomic sequencing tools such as environmental DNA (eDNA)
metabarcoding complement conventional fisheries approaches (e.g., netting) by providing in-depth, rapid, and non-invasive assessments of species composition. We implemented eDNA metabarcoding to assess its utility as an ecological monitoring tool in 5 estuarine sites in Kachemak Bay, Alaska with varying degrees of glacial coverage (0 – 60%). We analyzed (n=100) eDNA samples from April and September 2022 in Kachemak Bay to tease apart
seasonal and spatial patterns in fish diversity and to assess the extent to which environmental variables drive community composition. Results suggest a stronger influence of seasonal drivers on fish biodiversity than spatial drivers. Detections of fish taxa were higher using eDNA than
conventional beach seining. The results of this study will build on existing baseline data to document changes in a region heavily impacted by climate change.

A few things to note: 

All (seine and eDNA) data will be standardized from 0 to 1 to provide a semiquantitative value for comparison, this will be done with a Wisconsin double standardization. In this transformation, the abundance values are first standardized by species maximum standardization, and then by sample total standardization, and by convention multiplied by 100 (here we will have values from 0 to 1). Essentially, each element is divided by its column maximum and then divided by the row total. Bray and Curtis (1957) employed a double standardization before ordination. In their study, tree species were measured on different scales than were shrubs and herbs (density and basal area for trees, frequency for herbs and shrubs), so that a species maximum standardization achieved a common scale. Their rationalization for the subsequent sample total standardization was that not all samples had the same number of measurements, and that the stand total standardizations achieved a more uniform basis for comparison. The environmental variables will be 

standardized as follows: 
salinity-
temperature- 
DO- 
turbidity-


# First we will load the necessary libraries for this analysis.
```{r load libs}
# Install and load necessary packages
# if (!requireNamespace("vegan", quietly = TRUE)) install.packages("vegan")
 #if (!requireNamespace("ecodist", quietly = TRUE)) install.packages("ecodist")
# if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
# if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

library(vegan)
library(ecodist)
library(ggplot2)
library(tidyverse)
library(qiime2R)
library(phyloseq)
library(lubridate)
library(iNEXT)
library(ggVennDiagram)
library(fitdistrplus)
library(reshape2)
# devtools::install_github("donaldtmcknight/microDecon") #Installs microDecon
library(microDecon)
```

# Next we will load the data and take a look at it. 
```{r load data}
# Load the seine data including fish community data and taxa classification
seine <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/seine/2019_2020_2021_2022_Fish_Community_Data.csv")
seine_taxa <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/seine/fish_taxa_classification.csv")

# load in environmental data, this is also seine metadata
env <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/environmental/JNKB_20212022_Point_Sampling.csv")

# Load in the data from the L20230323_0629C Azenta run. This has samples from April and September 2022 in Kachemak Bay. 
edna_groups <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/eDNA/taxon_groups.csv")
edna_tax <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/eDNA/taxonomy_collapsed.csv")
method = "dada2-pe"
filtering = "unfiltered"
edna_table <- read_qza("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/eDNA/table.qza",method,filtering)

edna_meta <- read_tsv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/eDNA/metadata.tsv")

```
```{r reformatting}
# add in some reformatting of the data
edna_meta <- edna_meta %>%
  mutate(
    SamplingDate = mdy(SamplingDate),
    SamplingDate = format(SamplingDate, format = "%m/%d/%Y"),
    sample.name.library.prep = if_else(str_detect(sample_name, "NC$"),
                                        str_replace(sample_name, "NC$", "nc"),
                                        sample_name)
  )
# Clean invalid characters and then apply the trimming and replacing operations
edna_meta <- edna_meta %>%
  mutate(Region = iconv(Region, from = "UTF-8", to = "UTF-8", sub = ""),  # Remove invalid characters
         Region = str_trim(Region),         # Remove trailing spaces
         Region = str_replace_all(Region, "Kachemak Bay", "Kachemak Bay"))  # 

month_mapping <- c(
  "Jan" = "January", "February" = "February", "Feb" = "February",
  "Mar" = "March", "March" = "March", "Apr" = "April", "April" = "April",
  "May" = "May", "Jun" = "June", "June" = "June", "Jul" = "July",
  "July" = "July", "Aug" = "August", "August" = "August",
  "Sep" = "September", "September" = "September"
)

# Clean and standardize the SamplingMonth column
edna_meta <- edna_meta %>%
  mutate(SamplingMonth = str_trim(SamplingMonth),  # Remove leading/trailing spaces
         SamplingMonth = dplyr::recode(SamplingMonth, !!!month_mapping))  # Recode to consistent names

env <- env %>%
  mutate(SamplingDate = mdy(SamplingDate),
         SamplingDate = format(SamplingDate, format = "%m/%d/%Y"))
seine <- seine %>%
  mutate(SamplingDate = mdy(SamplingDate),
         SamplingDate = format(SamplingDate, format = "%m/%d/%Y"))
```

# Now we are going to get only the data we are interested in for this analysis which is the fish community data and the environmental data from Kachemak Bay in 2022 April and September. 
```{r kb data}
# subset the edna meta by the samples from Kachemak Bay in 2022 April and September
kb_edna_meta <- edna_meta %>%
  filter(SamplingYear.x == 2022, SamplingMonth %in% c("April", "September"), Region == "Kachemak Bay")

# now do the same for the seine data
kb_seine <- seine %>%
  filter(SamplingYear == 2022, SamplingPeriod %in% c("1", "6"), Region == "kb")
# now for the environmental data
kb_env <- env %>%
  filter(SamplingYear == 2022, SamplingMonth %in% c("April", "September"), region == "kb")
```

# Now we are going to standardize the seine data. 
```{r standardize}
# first let's create a site by species matrix from the kb_seine data, the count is in the column "Count", the species is in the column "ScientificName," and we will make a unique identifier for each sample in the column "SampleID" this will combine the SiteAcronym and the SamplingPeriod

# create the unique identifier
kb_seine$SampleID <- paste(kb_seine$SiteAcronym, kb_seine$SamplingPeriod, sep = "_")

# first we need to aggregate 
kb_seine <- kb_seine %>%
  group_by(SampleID, ScientificName) %>%
  summarise(Count = sum(Count)) %>% 
  ungroup()
# create the site by species matrix
site_by_species <- kb_seine %>%
  dplyr::select(SampleID, ScientificName, Count) %>%
  spread(key = ScientificName, value = Count, fill = 0)
seine_sampleID <- site_by_species$SampleID
# now make the rownames the SampleID and remove the column 
site_by_species <- site_by_species[, -1]
rownames(site_by_species) <- seine_sampleID

# let's look at our sample depth for the seine using the iNEXT package. First we need to transform the data so that it is in the correct format for the iNEXT package. This means transposing the data so that species are the rows and samples are the columns and making sure it is numeric. We will do that first. 
t_seine <- t(site_by_species)
t_seine <- site_by_species[,sapply(site_by_species, is.numeric)]
t_seine <- as.data.frame(t_seine)
t_seine <- na.omit(t_seine)
rownames(t_seine) <- seine_sampleID
# Remove species with zero counts across all samples
t_seine_filtered <- t_seine[,colSums(t_seine) > 0]

# Use the iNEXT package with the filtered data
seine_rare <- iNEXT(t_seine_filtered, datatype = "abundance")
plot(seine_rare)





# Create list for iNEXT Species Incidence Frequencies

# eventually we will want to compare the two sampling methods. So, for the sake of comparison, we will turn the seine data to presence absence
t_seine_filtered[t_seine_filtered > 0] <- 1
mor_inc <- list("Seine" = t(t_seine_filtered))

# Convert to incidence frequencies
species_incidence <- lapply(mor_inc, as.incfreq)

# Define the sequence for sample sizes
# t <- seq(1, 30, by=1)

# Run iNEXT, if we wanted to add sample size then add the argument size = t at the endof the function
out.inc2 <- iNEXT(species_incidence, q=0, datatype="incidence_freq")

# Plot the results
sample_coverage_fig <- ggiNEXT(out.inc2, type=2,color.var="Assemblage") + 
  theme_bw(base_size = 18) + 
  theme(plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 22, face = "bold"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        plot.subtitle = element_text(size = 18),
        legend.spacing.y = unit(0.5, 'cm'), 
        legend.key = element_rect(linewidth = 10),
        legend.key.size = unit(2, 'lines'))
# save the plot as a pdf
ggsave(plot = sample_coverage_fig, filename = "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_sample_coverage.pdf")

asymptotic_estimates <- out.inc2$DataInfo

# Interpretation: High Sample Coverage (SC = 0.8291): This suggests that the seine sampling effort has captured about 82.91% of the species diversity in the assemblage. While this is quite high, it also indicates that there are still some species that might be discovered with additional sampling.
```

```{r standardize}
# now we want to standardize the data with a wisconsin double standardization
wisconsin <- wisconsin(site_by_species) 

# this now gives us a standardized site by species matrix with the proportion of each species in each sample from 0 to 1 

# now we want to do the same for the environmental data, let's aggregate by site acronym and sampling date, we will take the mean of the environmental variables, if there are multiple samples for a site and date 
kb_env <- kb_env %>%
  group_by(site_id,SamplingMonth) %>%
  summarise(Salinity = mean(Salinity, na.rm = TRUE),
            Temperature = mean(Temperature, na.rm = TRUE),
            DissolvedOxygen = mean(DissolvedOxygen, na.rm = TRUE),
            #for turbidity we have Turbidity_rep1 and Turbidity_rep2 and Turbidity_rep3 we will take the mean of these
            Turbidity = mean(c(Turbidity_rep1, Turbidity_rep2, Turbidity_rep3), na.rm = TRUE),
  # add in error for each variable 
  Salinity_error = sd(Salinity, na.rm = TRUE),
         Temperature_error = sd(Temperature, na.rm = TRUE),
         DissolvedOxygen_error = sd(DissolvedOxygen, na.rm = TRUE),
         Turbidity_error = sd(c(Turbidity_rep1, Turbidity_rep2, Turbidity_rep3), na.rm = TRUE)) %>%
  ungroup()
# not getting error for anything except turbidity, will look at this later 
```

```{r}
# I just want to take a quick look at the seine data to see how many unique taxa we have 
unique_taxa <- kb_seine %>%
  dplyr::select(ScientificName) %>%
  distinct()
# alright so we have 27 unique taxa in the seine data 
# let's look at how many belong to the April samples and the September samples 
april_taxa <- seine %>%
  filter(SamplingPeriod == 1 & SamplingYear == 2022) %>%
  dplyr::select(ScientificName) %>%
  distinct()
# 16 unique taxa in april 
september_taxa <- seine %>%
  filter(SamplingPeriod == 6 & SamplingYear == 2022) %>%
  dplyr::select(ScientificName) %>%
  distinct()
# 26 unique taxa in september
# which were the taxa that were only found in september? and those that were only found in april? 
unique_september_taxa <- september_taxa %>%
  anti_join(april_taxa, by = "ScientificName")
print(unique_september_taxa) # 15 unique taxa to september 
unique_april_taxa <- april_taxa %>%
  anti_join(september_taxa, by = "ScientificName") # 5 unique taxa to april 
print(unique_april_taxa)
# so we have 5 unique taxa to april and 15 unique taxa to september, and 11 taxa that are shared between the two months. Let's print out the 11 that are shared 
shared_taxa <- april_taxa %>%
  inner_join(september_taxa, by = "ScientificName")
print(shared_taxa)

# Define the unique taxa found in April and September
# Extract the ScientificName column as vectors
april_taxa <- unique(april_taxa$ScientificName)
september_taxa <- unique(september_taxa$ScientificName)

# Create a list of the taxa
taxa_list <- list(April = april_taxa, September = september_taxa)

# Create the Venn diagram
ggVennDiagram(taxa_list) +
  ggtitle("Venn Diagram of Seine Samples")
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_venn_diagram.pdf")
# it is important to note that these are taxa and not species, in the seine data we have varying taxonomic levels, we can make a taxonomic classification for each of these taxa to see what level they were classified down to. I have to figure out how to join the seine data with the taxonomy data to do this. This might just be easiest to do manually, I am going to write out the unique taxa and then classify them manually. 
write.csv(unique_taxa, "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_unique_taxa.csv")

# read in the unique taxa with the classification
unique_taxa_class <- read.csv("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/seine/seine_unique_taxa_taxonomic_breakdown.csv")
# just a quick breakdown, of the unique taxa, how many were classified to the species level, genus level, family level, order level, class level, and phylum level. Let's first define the taxonomic level, this will be the column in the unique taxa class that has the classification. The order of rank is domain, kingdom, phylum, class, order, family, genus, species. I'll have R Read through each of the rows and classify the taxonomic level. If there is an NA it will write the column name in the taxonomic level column. If it goes all the way to species then it will write species in the taxonomic level column.
unique_taxa_class$TaxonomicLevel <- NA
for (i in 1:nrow(unique_taxa_class)){
  if (!is.na(unique_taxa_class$Species[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Species"
  } else if (!is.na(unique_taxa_class$Genus[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Genus"
  } else if (!is.na(unique_taxa_class$Family[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Family"
  } else if (!is.na(unique_taxa_class$Order[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Order"
  } else if (!is.na(unique_taxa_class$Class[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Class"
  } else if (!is.na(unique_taxa_class$Phylum[i])){
    unique_taxa_class$TaxonomicLevel[i] <- "Phylum"
  } else {
    unique_taxa_class$TaxonomicLevel[i] <- "NA"
  }
}

unique_taxa_class %>%
  group_by(TaxonomicLevel) %>%
  summarise(n = n())
# so for our 27 unique taxa, we have 23 identified down to the species level, 1 to the genus level, and three to the family level. 

# let's look at the unique taxa that were only found in september, what taxonomic level were they classified to?

unique_september_taxa_class <- unique_taxa_class %>%
  filter(ScientificName %in% unique_september_taxa$ScientificName)
print(unique_september_taxa_class)

unique_september_taxa_class %>%
  group_by(TaxonomicLevel) %>%
  summarise(n = n())
# 10 to species and 2 to family 
# now april 
unique_april_taxa_class <- unique_taxa_class %>%
  filter(ScientificName %in% unique_april_taxa$ScientificName)
print(unique_april_taxa_class)

unique_april_taxa_class %>%
  group_by(TaxonomicLevel) %>%
  summarise(n = n())
# three to species and 1 to family 




# let's plot out the taxonomy now for april and september at each site from our standardized data. 
# first we need to join the wisconsin data with the kb_seine data to get the taxonomy for each site
# separate seine meta by April and September of 2022 in kb 
kb_seine_meta <- seine %>%
  filter(SamplingPeriod %in% c("1", "6"), SamplingYear == 2022, Region == "kb")
# add in the sampling month if the Sampling period is 1 then the Sampling Month is April, if the Sampling Period is 6 then the Sampling Month is September
kb_seine_meta$SamplingMonth <- ifelse(kb_seine_meta$SamplingPeriod == 1, "April", "September")
kb_seine_meta$SampleID <- paste(kb_seine_meta$SiteAcronym, kb_seine_meta$SamplingPeriod, sep = "_")
# now we can join the kb_seine_meta with the wisconsin data after we aggregate the data by SampleID. We are only interested in the following columns, region, site_id, SamplingMonth, SampleID and SamplingPeriod

kb_seine_meta <- kb_seine_meta %>%
  dplyr::select(Region, SiteAcronym, SamplingMonth, SampleID, SamplingPeriod)
# aggregate the data by SampleID
kb_seine_meta <- kb_seine_meta %>%
  group_by(SampleID) %>%
  summarise(region = first(Region),
            site_id = first(SiteAcronym),
            SamplingMonth = first(SamplingMonth),
            SamplingPeriod = first(SamplingPeriod)) %>%
  ungroup()
# now we can join the kb_seine_meta with the wisconsin data
wisconsin <- as.data.frame(wisconsin)
# long format
wisconsin_long <- wisconsin %>%
  rownames_to_column(var = "SampleID") %>%
  gather(key = "ScientificName", value = "Proportion", -SampleID) %>%
  left_join(kb_seine_meta, by = "SampleID")
wisconsin_long <- wisconsin_long %>%  
  left_join(unique_taxa_class, by = c("ScientificName" = "ScientificName"))

# now we can plot the taxonomy for each site in April and September
ggplot(wisconsin_long, aes(x = site_id, y = Proportion, fill = ScientificName)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~SamplingMonth) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Fish Community Composition in Kachemak Bay 2022",
       x = "Taxon",
       y = "Proportion",
       fill = "Sampling Month")

# Let's see if we can plot based on the taxonomic level, this column tells us what taxonomic level the taxa was classified to. So I want to nest my legend by this column with ggnewscale.

# first we need to make a new column based on the taxonomic level column that will be the fill for the plot. This column tells us what level the taxa was identified to, so we need to use the column that has either "species," "genus," "family," or "NA" in it, then we need to use that value to fill in our new column that has the appropriate taxonomy based on that level and the taxa name from the other columns 
wisconsin_long$TaxonomicLevelName <- NA
for (i in 1:nrow(wisconsin_long)) {
  if (is.na(wisconsin_long$TaxonomicLevel[i])) {
    wisconsin_long$TaxonomicLevelName[i] <- "NA"
  } else if (wisconsin_long$TaxonomicLevel[i] == "Species") {
    wisconsin_long$TaxonomicLevelName[i] <- wisconsin_long$Species[i]
  } else if (wisconsin_long$TaxonomicLevel[i] == "Genus") {
    wisconsin_long$TaxonomicLevelName[i] <- wisconsin_long$Genus[i]
  } else if (wisconsin_long$TaxonomicLevel[i] == "Family") {
    wisconsin_long$TaxonomicLevelName[i] <- wisconsin_long$Family[i]
  } else {
    wisconsin_long$TaxonomicLevelName[i] <- "NA"
  }
}

# Install ggnewscale package if not already installed
if (!requireNamespace("ggnewscale", quietly = TRUE)) install.packages("ggnewscale")
library(ggnewscale)

# okay so now let's plot the taxonomy by taxonomic level
ggplot(wisconsin_long, aes(x = site_id, y = Proportion, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~SamplingMonth) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Seine Fish Community Composition in Kachemak Bay 2022",
       x = "Taxon",
       y = "Proportion",
       fill = "Taxonomic Level") +
  new_scale_fill() 
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_taxonomy_plot_wisconsin.png")
# this looks a bit funky, let's see what the data looks like without the wisconsin standardization
# join kb_seine with the kb_seine meta 
kb_seine <- kb_seine %>%
  left_join(kb_seine_meta, by = "SampleID")
# now join with taxonomic classification 
kb_seine <- kb_seine %>%
  left_join(unique_taxa_class, by = c("ScientificName" = "ScientificName"))
# now we can plot the taxonomy for each site in April and September
ggplot(kb_seine, aes(x = site_id, y = Count, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~SamplingMonth) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Fish Community Composition in Kachemak Bay 2022",
       x = "Taxon",
       y = "Count",
       fill = "Taxonomic Level")
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_taxonomy_plot_raw.png")
# okay we need to scale the data, let's do that now
kb_seine$Count <- scale(kb_seine$Count)

# now we can plot the taxonomy for each site in April and September
ggplot(kb_seine, aes(x = site_id, y = Count, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~SamplingMonth) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Fish Community Composition in Kachemak Bay 2022",
       x = "Taxon",
       y = "Count",
       fill = "Taxonomic Level")
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_taxonomy_plot_scaled.png")

```

```{r}
# let's do bray curtis dissimilarity and plot the results
# first we need to make a distance matrix
dist <- vegdist(wisconsin, method = "bray")

# now we can do a non-metric multidimensional scaling
nmds <- metaMDS(dist, k = 2, distance = "bray")

nmds_coords <- as.data.frame(nmds$points)
nmds_coords$SampleID <- site_by_species$SampleID # I had to do some fenagling for this, it will not run on its own until I fix the code above from deleting the sampleID column/rownames of the site_by species matrix

# add in a variable called Sampling Month which identifies the SampleID with the SamplingMonth. for example if the SampleID is jn_1 then the SamplingMonth is April
nmds_coords$SamplingMonth <- ifelse(str_detect(nmds_coords$SampleID, "1"), "April", "September")
# and site is the site acronym that comes before the underscore in the SampleID
nmds_coords$site <- str_extract(nmds_coords$SampleID, "^[^_]+")
# now we can plot the samples with ggplot by Sample ID
ggplot(nmds_coords, aes(x = MDS1, y = MDS2, color = SamplingMonth, shape = site)) +
  geom_point() +
  theme_minimal() +
  labs(title = "NMDS of Fish Community Data from Kachemak Bay 2022",
       x = "NMDS1",
       y = "NMDS2",
       color = "Sampling Month",
       shape = "Site")
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_NMDS_plot.png")

# Euclidean distance matrix
euclidean_dist <- dist(kb_env[,3:6], method = "euclidean")
# Mantel test
seine_mantel_test <- ecodist::mantel(dist ~ euclidean_dist)
print(seine_mantel_test)

# Correlation Coefficient (mantelr): The very low value (0.0083) indicates a very weak relationship between the biological and environmental distance matrices. P-values (pval1, pval2, pval3): All p-values are high (greater than 0.05), indicating that the observed correlation is not statistically significant. This means that there is no strong evidence to suggest a meaningful relationship between the two matrices.Confidence Interval (llim.2.5%, ulim.97.5%): The confidence interval includes zero, reinforcing the conclusion that the correlation is not significant. the Mantel test results suggest that there is no significant correlation between the biological community structure and the environmental variables in your data.

# PERMANOVA
permanova_test <- adonis2(wisconsin ~ SamplingMonth + site_id, data = kb_env, method = "bray")
print(permanova_test)

# this show that there is a significant difference in the fish community structure between the two sampling months, but not between the sites.


# okay, so I have averaged by site id and sampling month which is making it harder to interpret relationships between environmental variables and biological data. We do have multiple seines from a given site, so this is a way that we can add in a random effect to the model to account for this. this would  include a random effect for site in the analysis to account for the nested structure of the data.
permanova_test <- adonis2(wisconsin ~ SamplingMonth + Temperature + Salinity + Turbidity + DissolvedOxygen + (1|site_id), data = kb_env, method = "bray")

print(permanova_test)
```

Let's plot out the seine taxonomy data. 
```{r}
# we will use the wisconsin standardized data to plot out the taxonomy for each site 
# first we need to join the wisconsin data with the kb_seine data to get the taxonomy for each site
rownames(wisconsin) <- site_by_species$SampleID

```

# eDNA data analysis. We are going to start with the data that has not been decontaminated following Zack's protocol. If it seems like there is some bizarre stuff going on with the data we will go back and decontaminate the data. 
# Okay now we are going to decontam. 

```{r}
# let's prep the hash key for when we might decontaminate, this is also gonna help us build our ASV table 
tax_table <- edna_tax %>% 
  mutate(sum_taxa = paste(kingdom, phylum, class, order, family, genus, species, sep = ",")) %>%
    # Rename 'Feature.ID' to 'Hash'
    rename(Hash = qseqid)

Hash_key_blast <- tax_table

Hash_key_blast %>% 
  distinct(Hash, .keep_all = T) -> Hash_key_blast

process_count_table <- function(file_path) {
  # Read the count table from the .qza file
  count_table <- read_qza(file = file_path)$data
  
  # Convert it to a data frame
  count_table <- as.data.frame(count_table)
  
  # Convert row names (Hashes) into a column called "Hash"
  count_table_df <- rownames_to_column(count_table, var = "Hash")
  
  # Reshape the data from wide to long format
  count_table_long <- count_table_df %>%
    pivot_longer(cols = -Hash, names_to = "sample_id", values_to = "nReads")
  
  # Return the transformed table
  return(count_table_long)
}

file_path <- "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/inputs/eDNA/table.qza"

# Process the count table
ASV_table <- process_count_table(file_path)
# now join the ASV_table with the tax_table
ASV_table <- ASV_table %>%
  left_join(Hash_key_blast, by = "Hash")
# now we need to subset the data to only include the samples from Kachemak Bay in 2022 April and September
# now we can join the ASV_table with the kb_edna_meta
ASV_table <- ASV_table %>%
  inner_join(edna_meta, by = c("sample_id" = "sample_name"))

# Fill in the ASV table with sample type and join with the cleaned hash key 
ASV_table %>%
  mutate(source = case_when(
    str_detect(sample_id, "eblank") ~ "Extraction NC",
    str_detect(sample_id, "tilapia|mahi") ~ "Positives",
    str_detect(sample_id, "technc") ~ "Negatives",
    str_detect(sample_id, "nc") & !str_detect(sample_id, "technc") ~ "Field Negatives",
    TRUE ~ "Samples"
  )) -> ASV_table
ASV_table %>% 
  inner_join(Hash_key_blast, by=c("Hash"="Hash")) -> ASV_table_blast 
ASV_table_blast %>%
  dplyr::group_by(sum_taxa.x) %>%
  mutate (TotalReadsperSample = sum(nReads)) %>% 
  filter(., TotalReadsperSample > 1) %>% 
  dplyr::select(-TotalReadsperSample) -> ASV_table_used
ASV_table_blast %>%  dim() -> all_dim
ASV_table_used %>%  dim() -> used_only_dim
paste0(round(used_only_dim[[1]]/all_dim[[1]]*100,2),"% ASVs retained")#100% ASVs retained"
kb_ASV_table <- ASV_table_used %>%
  filter((SamplingYear.x == 2022 & 
          SamplingMonth %in% c("April", "September") & 
          Region == "Kachemak Bay" & 
          source == "Samples" & 
          !grepl("^lc", sample_id)) | 
         (source %in% c("Positives", "Negatives", "Field Negatives") & 
          !grepl("^lc", sample_id)))
kb_ASV_table$SampleID <- paste(kb_ASV_table$Site, kb_ASV_table$SamplingMonth, sep = "_")

```
```{r}
# Identify blank samples
blank_samples <- kb_ASV_table %>%
  filter(source == "Field Negatives")
numb.blanks <- nrow(blank_samples)

# Add blank samples
step.3_barcode1 <- bind_rows(blank_samples, kb_ASV_table)

kb_ASV_table_summarized <- kb_ASV_table %>%
  group_by(Hash, sample_id) %>%
  summarise(nReads = sum(nReads), .groups = "drop")

# Pivot to wide format
step.3_barcode1_wide <- kb_ASV_table_summarized %>%
  pivot_wider(names_from = sample_id, values_from = nReads, values_fill = list(nReads = 0))

# Assuming metadata is available and has the correct columns
metadata <- kb_ASV_table %>% 
  dplyr::select(sample_id, Site, SamplingMonth) %>% 
  distinct()

# Assign groups
sample_counter <- step.3_barcode1_wide %>%
  pivot_longer(cols = -Hash, names_to = "sample_id", values_to = "nReads") %>%
  left_join(metadata, by = "sample_id") %>%
  group_by(Site, SamplingMonth) %>%
  count()
numb.ind <- sample_counter$n

step.3_barcode1_decon <- decon(
   data = step.3_barcode1_wide,
   numb.blanks = nrow(blank_samples),
   numb.ind = sample_counter$n,
   taxa = TRUE,
   runs = step3_runs,
   thresh = step3_thresh,
   prop.thresh = step3_prop.thresh,
   regression = step3_regression
 )
```


```{r}
# Assuming your data frame is named kb_ASV_table

# Process the data and create the bar plot
kb_ASV_table %>%
  mutate(nReads = as.numeric(nReads)) %>%
  mutate(nReads = ifelse(is.na(nReads), 0, nReads)) %>%
  filter(source == "Positives") %>%
  filter(nReads > 0) %>%
  ggplot(aes(x = sample_id, y = nReads, fill = sum_taxa.x)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  labs(
    y = "number of sequencing reads",
    x = "sample ID",
    title = "ASV reads - positive controls"
  ) + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    #legend.position = "none",
    legend.title = element_blank()
  )
# Calculate the maximum proportion of reads for each ASV in positive controls:

prop_asvs_in_positives <- kb_ASV_table %>%
  filter(source == "Positives") %>%
  group_by(sample_id) %>%
  mutate(TotalReadsPerSample = sum(nReads)) %>%
  mutate(Prop = nReads / TotalReadsPerSample) %>%
  group_by(Hash) %>%
  summarise(max_prop = max(Prop)) %>%
  arrange(desc(max_prop))
# Subtract the maximum proportion of tag-jumped reads for each ASV from samples:

indexhop_table <- kb_ASV_table %>%
  mutate(nReads = as.numeric(nReads)) %>%
  mutate(nReads = ifelse(is.na(nReads), 0, nReads)) %>%
  group_by(sample_id) %>%
  mutate(TotalReadsPerSample = sum(nReads, na.rm = TRUE)) %>%
  left_join(prop_asvs_in_positives, by = c("Hash")) %>%
  mutate(IndexHoppingReads = TotalReadsPerSample * max_prop) %>%
  mutate(reads_IndexHop_removed = nReads - IndexHoppingReads) %>%
  mutate(reads_IndexHop_removed = if_else(reads_IndexHop_removed < 0, 0, reads_IndexHop_removed))

# Clean up the table by removing columns no longer needed:
asv_table_filter1 <- indexhop_table %>%
  dplyr::select(sample_id, source, Hash, reads_IndexHop_removed) %>%
  dplyr::rename(reads = reads_IndexHop_removed)
decontaminated_1 <- indexhop_table %>%
  dplyr::select(sample_id, Hash, IndexHoppingReads) %>%
  pivot_wider(names_from = "Hash", values_from = "IndexHoppingReads")


reads_per_type_ASV <- asv_table_filter1 %>%
  group_by(Hash, source) %>%
  summarize(TotalReadsPerASV = sum(reads, na.rm = TRUE)) %>%
  arrange(Hash)
```
what ASVs have no reads in samples, but reads in the controls? 
```{r}
not_in_samples <- reads_per_type_ASV %>%
  pivot_wider(names_from = "source", values_from = c("TotalReadsPerASV")) %>%
    filter(Samples < 1 & Positives > 0 & Negatives > 0 & `Field Negatives` > 0)
not_in_samples
```
```{r}
asv_table_wide <- asv_table_filter1 %>%
  dplyr::select(!source) %>%
  mutate(reads = as.integer(reads)) %>%
  pivot_wider(names_from = Hash, values_from = reads)

sample_IDs <- asv_table_wide$sample_id

asv_table_wide <- asv_table_wide %>%
  ungroup() %>%
  dplyr::select(-sample_id)

## plots the figure
rarecurve(asv_table_wide, step = 20, col = "blue", label = FALSE, 
          main = "Sequencing Effort Curves",
          xlab = "Sequencing Depth", ylab = "Number of ASVs Identified",
          xlim = c(0,5000))
```

summarize in a table how many pcr replicates meet certain read count thresholds 
```{r}
read_summary <- asv_table_filter1 %>%
  group_by(sample_id, source) %>%
  summarize(tot_reads = sum(reads)) %>%
  arrange(desc(tot_reads)) %>%
  group_by(source) %>%
  summarize(atleast1 = sum(tot_reads >= 1),
            atleast250 = sum(tot_reads >= 250),
            atleast500 = sum(tot_reads >= 500),
            atleast750 = sum(tot_reads >= 750),
            atleast1k = sum(tot_reads >= 1000),
            atleast2k = sum(tot_reads >= 2000))
read_summary
```

based on taxa accumulation curve and summary table, we will remove any pcr replicate with fewer than 1000 reads from downstream analyses

```{r}
reps_below <- asv_table_filter1 %>%
  group_by(sample_id) %>%
  summarise(tot_reads = sum(reads)) %>%
  filter(tot_reads < 1000)
asv_table_filter <- asv_table_filter1 %>%
  filter(!sample_id %in% reps_below$sample_id)
asv_table_filtered <- asv_table_filter %>%
  inner_join(kb_ASV_table, by = "Hash", relationship = "many-to-many") %>%
  filter(!(genus.x %in% c("Coryphaena","Oreochromis")))
write.csv(asv_table_filtered, "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/ASV_table_filtered.csv")
```

```{r}
# Step 1: Identify negative control hashes
negative_control_hashes <- kb_ASV_table %>%
  filter(source == "Negatives" & nReads > 0) %>%
  pull(Hash) %>%
  unique()

# Step 2: Filter the data for negative control hashes
filtered_data <- kb_ASV_table %>%
  filter(Hash %in% negative_control_hashes)

# Step 3: Calculate the total reads per sample
filtered_data <- filtered_data %>%
  group_by(sample_id) %>%
  mutate(total_reads_per_sample = sum(nReads)) %>%
  ungroup()

# Step 4: Calculate the read percentage
filtered_data <- filtered_data %>%
  mutate(read_percentage = nReads / total_reads_per_sample)

# Step 5: Define the threshold percentage (example value, replace with actual threshold)
threshold_percentage <- 0.015

# Step 6: Filter based on the threshold percentage
filtered_data <- filtered_data %>%
  filter(read_percentage >= threshold_percentage)

# Print the filtered data
print(filtered_data)
```

```{r}
# Calculate negative control reads for the most abundant fish species
negative_control_reads <- filtered_data %>%
  filter(source == "Negatives") %>%
  group_by(Hash) %>%
  summarise(total_reads = sum(nReads)) %>%
  arrange(desc(total_reads)) %>%
  slice(1) %>%
  pull(total_reads)

# Calculate the number of negative controls
number_of_negative_controls <- kb_ASV_table %>%
  filter(source == "Negatives") %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n)

# Calculate the number of water samples
number_of_water_samples <- kb_ASV_table %>%
  filter(source == "Samples") %>%
  summarise(n = n_distinct(sample_id)) %>%
  pull(n)

# Calculate the total target fish reads
total_target_fish_reads <- kb_ASV_table %>%
  filter(source == "Samples") %>%
  summarise(total_reads = sum(nReads)) %>%
  pull(total_reads)

# Print the calculated values
negative_control_reads
number_of_negative_controls
number_of_water_samples
total_target_fish_reads

per_species_per_control_contamination <- negative_control_reads / number_of_negative_controls
total_potential_per_taxa_contamination <- per_species_per_control_contamination * number_of_water_samples
threshold_cutoff <- total_potential_per_taxa_contamination / total_target_fish_reads

# Step 2: Filter out species that contribute equal to or less than 0.55% of the total target fish reads in a sample
filtered_data <- kb_ASV_table %>%
  group_by(sample_id) %>%
  mutate(total_reads_per_sample = sum(nReads)) %>%
  ungroup() %>%
  mutate(read_percentage = nReads / total_reads_per_sample) %>%
  filter(read_percentage > threshold_cutoff)

# Step 3: Inspect sample level rarefaction curves
# Prepare data for rarefaction curves
rarefaction_data <- filtered_data %>%
  select(sample_id, Hash, nReads) %>%
  spread(key = Hash, value = nReads, fill = 0)

# Plot rarefaction curves
plot_rarefaction_curves <- function(data) {
  rarefaction_data <- rarefy(data, sample = min(rowSums(data)))
  rarecurve(rarefaction_data, step = 20, sample = min(rowSums(data)), col = "blue", cex = 0.6)
}

plot_rarefaction_curves(rarefaction_data)

# Step 4: Check the assemblage composition using nMDS
# Prepare data for nMDS
nmds_data <- filtered_data %>%
  select(sample_id, Hash, nReads) %>%
  spread(key = Hash, value = nReads, fill = 0) %>%
  column_to_rownames(var = "sample_id")

# Perform nMDS
nmds_result <- metaMDS(nmds_data, distance = "bray", k = 2, trymax = 100)

# Plot nMDS
plot(nmds_result, type = "t")

# Check for outliers
stressplot(nmds_result)
```
### Step 1: Nest the dataset and split it in positives and samples
```{r}
kb_ASV_table %>% 
  group_by(source) %>% 
  nest() %>% 
  pivot_wider(names_from = source, values_from =  data) -> ASV_nested 
how.many <- function(ASVtable, round){
  ASVtable %>% ungroup() %>% 
    summarise(nsamples = n_distinct(sample_id),
              nHashes = n_distinct(Hash),
              nReads = sum(nReads), 
              Stage = paste0("Step_", round)) %>% 
    gather(starts_with("n"), value = "number", key = "Stat")
}

ASV_nested %>% 
  transmute(Summary = map(Samples, ~ how.many(ASVtable = .,round = 0)))  -> ASV_summary

ASV_summary %>% 
  unnest(Summary)
```
### Step 2: Model the composition of the positive controls of each run 
We create a vector of the composition of each positive control and substract it from the environmental samples from their runs
```{r}
ASV_nested %>% 
  mutate (contam.tibble = map(Positives, 
                              function(.x){
                                .x %>%
                                  group_by(sample_id) %>%
                                  mutate (TotalReadsperSample = sum(nReads)) %>%
                                  mutate (proportion = nReads/TotalReadsperSample) %>%
                                  group_by (Hash) %>%
                                  summarise (vector_contamination = max (proportion))
                                }) ) -> ASV_nested
ASV_nested %>% 
  dplyr::select(contam.tibble) %>% 
  unnest(cols = contam.tibble) # Check how it looks like
ASV_nested$contam.tibble[[1]] %>% as.data.frame() %>% 
  ggplot(aes(x= vector_contamination))+
  geom_histogram() -> vc_plot_1
# ggsave the vector contamination plot
ggsave(vc_plot_1, filename = "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/vector_contamination_plot.pdf")
```
### Step 3: Substract the composition of the positive controls from the environment samples

The idea behind this procedure is that we know, for each run, how many reads from each Hash appeared in the positive controls. These come from 2 processes: sequences we know should appear in the positive controls, and sequences that have *jumped* from the environment to the positive controls. With this procedure, we substract from every environmental sample the proportion of reads that jumped from elsewhere.
```{r}
ASV_nested %>% 
  mutate(cleaned.tibble = map2(Samples, contam.tibble, function(.x,.y){ 
    .x %>%
      group_by (sample_id) %>%
      mutate (TotalReadsperSample = sum (nReads)) %>%
      left_join(.y, by = "Hash") %>%
      mutate (Updated_nReads = ifelse (!is.na(vector_contamination),  nReads - (ceiling(vector_contamination*TotalReadsperSample)), nReads)) %>%
      filter (Updated_nReads > 0) %>%
      ungroup() %>% 
      dplyr::select (sample_id, Hash, nReads = Updated_nReads)
      
    
  })) -> ASV_nested
ASV_nested %>% 
  dplyr::select(cleaned.tibble) %>% 
  unnest(cleaned.tibble) #Check how they look

```

```{r summary.file.2}
# Add a common column to ASV_summary
ASV_summary <- ASV_summary %>%
  mutate(common_column = 1)

# Create the new summary data frame with the common column
ASV_nested %>% 
  transmute(Summary.1 = map(cleaned.tibble, ~ how.many(ASVtable = ., round = "1.Jump"))) %>% 
  mutate(common_column = 1) %>% 
  left_join(ASV_summary, by = "common_column") %>% 
  mutate(Summary = map2(Summary, Summary.1, bind_rows)) %>%
  dplyr::select(-Summary.1, -common_column) -> ASV_summary

ASV_summary %>% unnest(Summary)
```

## Cleaning Process 2: **Discarding samples with low number of reads**

We will fit the number of reads assigned to each sample to a normal distribution and discard those samples with a probability of 95% of not fitting in that distribution. The output would be a dataset with less samples and potentially less number of unique Hashes.

We are splitting this step up by sequencing run because we have a 10-fold difference in the output of reads for each sequecning run (this was because we were taking advantage of an Azenta deal)
```{r fitting nReads per sample}
###Pull out Sample Read Depth

ASV_nested %>% 
  dplyr::select(cleaned.tibble) %>% 
  unnest(cleaned.tibble) %>% 
  group_by(sample_id) %>%
  dplyr::summarise(tot = sum(nReads)) %>% 
  arrange(desc(tot))-> all_reps

all_reps %>% 
  pull(tot) -> reads.per.sample

names(reads.per.sample) <- all_reps %>% pull(sample_id)  

### Fit Normal Distribution
low_dist_probability_cutoff <- 0.025
minimum_read_cutoff <- 1000
high_dist_probability_cutoff <- 1

fit_1 <- fitdist(reads.per.sample, "gamma", lower=c(0,0), start=list(scale=1,shape=1))

all_reps %>%  
  mutate(prob = pgamma(tot, shape = fit_1$estimate[[2]], scale = fit_1$estimate[[1]], lower.tail = TRUE,
                       log.p = FALSE)) -> all_reps_1

### Remove Outlier Samples
outliers_1 <- all_reps_1 %>% 
  filter(prob < low_dist_probability_cutoff  | tot < minimum_read_cutoff | prob > high_dist_probability_cutoff) # changed to 0.025, this saved 4 samples, cutoff was originally 0.05
outliers_1 %>% View()

ASV_nested %>% 
  mutate(Step.1.low.reads = map (cleaned.tibble, ~ filter(.,!sample_id %in% outliers_1$sample_id) %>% ungroup)) -> ASV_nested
# Add a common column to ASV_summary
ASV_summary <- ASV_summary %>%
  mutate(common_column = 1)

# Create the new summary data frame with the common column
ASV_nested %>% 
  transmute(Summary.1 = map(Step.1.low.reads, ~ how.many(ASVtable = ., round = "2.Low.nReads"))) %>% 
  mutate(common_column = 1) %>%  # Add the common column here
  left_join(ASV_summary, by = "common_column") %>% 
  mutate(Summary = map2(Summary, Summary.1, bind_rows)) %>%
  dplyr::select(-Summary.1, -common_column) -> ASV_summary
ASV_summary %>% unnest(Summary)



```

## Cleaning Process 3: **Full clearance from Positive control influence**

Removing the Hashes that belong to the positive controls. First, for each Hash that appeared in the positive controls, determine whether a sequence is a true positive or a true environment. For each Hash, we will calculate, maximum, mean and total number of reads in both positive and samples, and then we will use the following decision tree:

  * If all three statistics are higher in one of the groups, we will label it either of Environmental or Positive control influence.
  
  * If there are conflicting results, we will use the Hashes. to see if they belong to either the maximum abundance of a Hash is in a positive, then it is a positive, otherwise is a real sequence from the environment.


Now, for each Hash in each set of positives controls, calculate the proportion of reads that were missasigned - they appeared somewhere they were not expected.
We will divide that process in two: first . A second step would be to create a column named proportion switched, which states the proportion of reads from one Hash that jumped from the environment to a positive control or viceversa. The idea is that any presence below a threshold can be arguably belong to tag jumping.
```{r}
kb_ASV_table %>% 
  filter (source == "Positives") %>% 
  group_by(Hash) %>% 
  summarise(tot = sum(nReads)) %>% 
  arrange(desc(tot)) %>% 
  pull(Hash) -> good.order
kb_ASV_table %>% 
  filter (Hash %in% good.order) %>%
  group_by(sample_id) %>% 
  mutate(tot.reads = sum(nReads)) %>% 
  group_by(Hash,sample_id) %>% 
  mutate(prop = nReads/tot.reads) %>% 
  group_by(Hash, source) %>% 
  summarise (max.  = max(prop),
             mean. = mean(prop),
             tot.  = sum(nReads)) %>% 
  gather(contains("."), value = "number", key = "Stat") %>%
  spread(key = "source", value = "number", fill = 0) %>% 
  group_by(Hash, Stat) %>%
  mutate(origin = case_when(Positives > Samples ~ "Positive.control",
                            TRUE                ~ "Environment")) %>% 
  group_by (Hash) %>%
  mutate(tot = n_distinct(origin)) -> Hash.fate.step2
Hash.fate.step2 %>% 
  filter(tot == 1) %>% 
  group_by(Hash) %>% 
  summarise(origin = unique(origin)) %>% 
  filter(origin == "Positive.control") -> Hashes.to.remove.step2

Hashes.to.remove.step2 %>% 
  left_join(Hash_key_blast, by = "Hash") -> Hashes.to.remove.step2

```
### Remove the positive control hashes from the composition of the ASVs

```{r cleaning.Step2}
ASV_nested %>% 
  mutate(Step2.tibble = map (Step.1.low.reads, ~ filter(.,!Hash %in% Hashes.to.remove.step2$Hash) %>% ungroup)) -> ASV_nested
#save RDS
#saveRDS(ASV_nested, file = "/Users/marisgoodwin/Documents/GITHUB/eDNA_estuaries/Output/ASV_nested.rds")
ASV_summary <- ASV_summary %>%
  mutate(common_column = 1)
# Create the new summary data frame with the common column

ASV_nested %>% 
  transmute( Summary.1 = map(Step2.tibble, ~ how.many(ASVtable = .,round = "3.Positives"))) %>% 
  mutate(common_column = 1) %>%
  left_join(ASV_summary, by = "common_column") %>% 
  mutate(Summary   = map2(Summary, Summary.1, bind_rows)) %>%
  dplyr::select(-Summary.1, -common_column) -> ASV_summary 
ASV_summary %>% 
  unnest()
```

# Cleaning Process 4: **Dissimilarity between PCR (biological) replicates** 

```{r dissimilarity between PCR replicates}
ASV_nested %>% 
  dplyr::select(Step2.tibble) %>% 
  unnest(Step2.tibble) %>%
  separate(sample_id, into = "original_sample", sep = "\\.", remove = F) -> cleaned.tibble
```
```{r quick check}
# do all samples have a name
cleaned.tibble %>% 
  filter (sample_id == "")
# do all of them have an original sample
cleaned.tibble %>% 
  filter(original_sample == "")
# do all of them have a Hash
cleaned.tibble %>% 
  filter(is.na(Hash))
# How many samples, how many Hashes
cleaned.tibble %>%
  ungroup %>% 
  summarise(n_distinct(sample_id), # 86
            n_distinct(Hash))   # 667

# Let's check the levels of replication
tibble <- cleaned.tibble %>%
  # Remove samples with 'eblank' in the name
  filter(!str_detect(sample_id, "eblank")) %>%
  # Extract 'rep' and 'biol' from the sample_id
  mutate(
    PCR = str_sub(sample_id, -1),  # Last character
    biol = str_sub(sample_id, -2, -2)  # Second-to-last character
  ) %>%
  # Split the remaining part of the sample_id to extract 'Site'
  mutate(Site = str_sub(sample_id, 1, -4)) %>%
  # Group by Site and nest the data
  group_by(Site) 

tibble <- tibble %>%
  mutate(bio_rep = str_remove(sample_id, ".$"))


tibble %>% 
  group_by(Site) %>% 
  summarise(nrep = n_distinct(PCR)) %>% 
  #filter (nrep == 3) # all the rest
  #filter (nrep == 2) # 13 
 filter (nrep == 1) # 3 
```

```{r remove single replicates}
discard.1 <- tibble %>% 
  group_by(Site) %>% 
  mutate(nrep = n_distinct(PCR)) %>% 
  #filter (nrep == 2) # 25
  filter (nrep == 1) %>% 
  distinct(Site) %>% pull(Site)

tibble %>% 
  filter(!Site %in% discard.1) -> cleaned.tibble
```

```{r lets do the PCR replication}
cleaned.tibble <- cleaned.tibble %>%  
  inner_join(Hash_key_blast, by=c("Hash"="Hash"))
cleaned.tibble %>%
  group_by(sample_id) %>%
  mutate(Tot = sum(nReads),
          Row.sums = nReads / Tot) %>% 
  group_by(sum_taxa) %>%
  mutate(Colmax = max(Row.sums,na.rm = T),
          Normalized.reads = Row.sums / Colmax) -> cleaned.tibble

tibble_to_matrix_sum_taxa <- function (tb) {
  
  tb %>% 
    group_by(sample_id, sum_taxa) %>% 
    summarise(nReads = sum(Normalized.reads)) %>% 
    spread(key = "sum_taxa", value = "nReads", fill = 0) -> matrix_1
    samples <- pull (matrix_1, sample_id)
    matrix_1 %>% 
      ungroup() %>% 
    dplyr::select( - sample_id) -> matrix_1
    data.matrix(matrix_1) -> matrix_1
    dimnames(matrix_1)[[1]] <- samples
    vegdist(matrix_1, method = "jaccard") -> matrix_1
}


tibble_to_matrix_sum_taxa(cleaned.tibble) -> all.distances.full
# Do all samples have a name?
summary(is.na(names(all.distances.full))) # Yes they do
```
make the pairwise distances a long table
```{r}
as_tibble(subset(melt(as.matrix(all.distances.full)))) -> all.distances.melted

# Any mjor screw ups
summary(is.na(all.distances.melted$value))

all.distances.to.plot <- all.distances.melted %>%
  left_join(edna_meta, by = c("Var1" = "sample_name"), relationship = "many-to-many") %>%
  rename(region_var1 = Region, site_var1 = Site, bottle_var1 = bottle.number, sampling_date1 = sampling.date.code) %>%
  left_join(edna_meta, by = c("Var2" = "sample_name"), relationship = "many-to-many") %>%
  rename(region_var2 = Region, site_var2 = Site, bottle_var2 = bottle.number, sampling_date2 = sampling.date.code) %>%
  dplyr::select(Var1, Var2, value, region_var1, site_var1, bottle_var1, region_var2, site_var2, bottle_var2, sampling_date1, sampling_date2) %>%
  
  # Extract technical replicate for Var1
  mutate(tech_rep1 = case_when(
    str_detect(Var1, "NC$") ~ "NC",  # If sample ends with "NC", set tech_rep to "NC"
    TRUE ~ str_extract(Var1, "\\d(?=\\D*$)")
  )) %>%
  
  # Extract technical replicate for Var2
  mutate(tech_rep2 = case_when(
    str_detect(Var2, "NC$") ~ "NC",  # If sample ends with "NC", set tech_rep to "NC"
    TRUE ~ str_extract(Var2, "\\d(?=\\D*$)")
  )) %>%
  
  # Create additional columns to compare sample types
  mutate(
    Distance.type = case_when(
      bottle_var1 == bottle_var2 & site_var1 == site_var2 & sampling_date1 == sampling_date2  ~ "PCR.replicates",  # Same PCR replicates
      site_var1 == site_var2 & sampling_date1 == sampling_date2 ~ "Biol.replicates",          # Same biological replicates
      region_var1 == region_var2 & site_var1 == site_var2 ~ "Same Region Same Site",          # Same region and same site
      region_var1 == region_var2 & site_var1 != site_var2 ~ "Same Region Different Site",     # Same region, different sites
      TRUE ~ "Different"                                                                      # Different region and site
    )
  ) %>%
  
  # Select and rename columns for the final output
  dplyr::select(Sample1 = Var1, Sample2 = Var2, value, Distance.type) %>%
  
  # Exclude comparisons of the same sample
  filter(Sample1 != Sample2)



# Checking all went well

sapply(all.distances.to.plot, function(x) summary(is.na(x)))

all.distances.to.plot$Distance.type <- all.distances.to.plot$Distance.type  %>% fct_relevel( "PCR.replicates", "Biol.replicates", "Same Region Different Site","Same Region Same Site")

  ggplot (all.distances.to.plot ) +
  geom_histogram (aes (fill = Distance.type, x = value, after_stat(ndensity)), position = "dodge",  alpha = 0.9, bins = 50) +
  facet_wrap( ~ Distance.type) +
  labs (x = "Pairwise dissimilarity", y = "density" ,
        Distance.type = "Distance") +
    guides (fill = "none")
ggsave("visual.anova.png", dpi = "retina")
```

```{r}
kb_ASV_table <- kb_ASV_table %>%
  group_by(SampleID, sum_taxa.x) %>%
  mutate(Count = sum(nReads)) %>%
  ungroup() %>%
  distinct()

# create the site by species matrix
site_by_species_eDNA <- kb_ASV_table %>%
  dplyr::select(SampleID, sum_taxa.x, Count) %>%
  spread(key = sum_taxa.x, value = Count, fill = 0)
edna_sampleID <- site_by_species_eDNA$SampleID
site_by_species_eDNA <- site_by_species_eDNA[, -1]

rownames(site_by_species_eDNA) <- edna_sampleID

t_edna <- t(site_by_species_eDNA)
t_edna <- site_by_species_eDNA[,sapply(site_by_species_eDNA, is.numeric)]
t_edna <- as.data.frame(t_edna)
t_edna <- na.omit(t_edna)
rownames(t_edna) <- edna_sampleID
# Remove species with zero counts across all samples
t_edna_filtered <- t_edna[,colSums(t_edna) > 0]

# Create list for iNEXT Species Incidence Frequencies

# eventually we will want to compare the two sampling methods. So, for the sake of comparison, we will turn the seine data to presence absence
t_edna_filtered[t_edna_filtered > 0] <- 1

mor_inc <- list("Seine" = t(t_seine_filtered),
                "eDNA" = t(t_edna_filtered))

# Convert to incidence frequencies
species_incidence <- lapply(mor_inc, as.incfreq)

# Define the sequence for sample sizes
# t <- seq(1, 30, by=1)

# Run iNEXT, if we wanted to add sample size then add the argument size = t at the endof the function
out.inc2 <- iNEXT(species_incidence, q=0, datatype="incidence_freq")

# Plot the results
sample_coverage_fig <- ggiNEXT(out.inc2, type=2,color.var="Assemblage") + 
  theme_bw(base_size = 18) + 
  theme(plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 22, face = "bold"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        plot.subtitle = element_text(size = 18),
        legend.spacing.y = unit(0.5, 'cm'), 
        legend.key = element_rect(linewidth = 10),
        legend.key.size = unit(2, 'lines'))

# save the plot as a pdf
ggsave(plot = sample_coverage_fig, filename = "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/eDNA_seine_sample_coverage.pdf")
asymptotic_estimates <- out.inc2$DataInfo
# Seine Assemblage: With a sample coverage of 82.91%, the seine observed most of the species, but additional sampling could have revealed more species. The majority of species are found in only one or two sites, indicating a relatively high level of site-specific species.

# on the other hand, the eDNA assemblage has a sample coverage of 97.62%, indicating that nearly all the species diversity has been captured. The species are more evenly distributed across the sites, with several species found in multiple sites, indicating a more homogeneous distribution. 


```

Let's look at the taxonomy of the eDNA data. 
```{r}
# Filter and select unique taxa for April and September with Count > 0
april_taxa_edna <- kb_ASV_table %>%
  filter(grepl("April", SampleID) & Count > 0) %>%
  dplyr::select(sum_taxa.x) %>%
  distinct() %>%
  pull(sum_taxa.x)

september_taxa_edna <- kb_ASV_table %>%
  filter(grepl("September", SampleID) & Count > 0) %>%
  dplyr::select(sum_taxa.x) %>%
  distinct() %>%
  pull(sum_taxa.x)

# Create a list of the taxa
taxa_list <- list(April = april_taxa_edna, September = september_taxa_edna)

# Create the Venn diagram with labels
venn_plot <- ggVennDiagram(taxa_list, label_alpha = 0) +
  ggtitle("Venn Diagram of eDNA samples") +
  theme(legend.position = "none") # Remove the legend for clarity
ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/eDNA_venn_diagram.pdf")
# Extract the unique and shared taxa
unique_april <- setdiff(april_taxa_edna, september_taxa_edna)
unique_september <- setdiff(september_taxa_edna, april_taxa_edna)
shared_taxa <- intersect(april_taxa_edna, september_taxa_edna)

# Add labels for unique and shared taxa
write.csv(kb_ASV_table, "/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/eDNA_ASV_table.csv")

# Display the Venn diagram
print(venn_plot)

kb_ASV_table$TaxonomicLevel <- NA
for (i in 1:nrow(kb_ASV_table)){
  if (!is.na(kb_ASV_table$species.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Species"
  } else if (!is.na(kb_ASV_table$genus.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Genus"
  } else if (!is.na(kb_ASV_table$family.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Family"
  } else if (!is.na(kb_ASV_table$order.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Order"
  } else if (!is.na(kb_ASV_table$class.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Class"
  } else if (!is.na(kb_ASV_table$phylum.x[i])){
    kb_ASV_table$TaxonomicLevel[i] <- "Phylum"
  } else {
    kb_ASV_table$TaxonomicLevel[i] <- "NA"
  }
}

# Group by TaxonomicLevel and sum_taxa, then count the occurrences
taxonomic_counts <- kb_ASV_table %>%
  group_by(TaxonomicLevel, sum_taxa.x, nReads > 0) %>%
  summarise(n = n()) %>%
  distinct()

# Summarize the counts for each taxonomic level
taxonomic_counts %>%
     group_by(TaxonomicLevel,`nReads > 0` == "TRUE") %>%
     summarise(total_count = n())



```
Now compare eDNA and seine 
```{r}
# Extract unique taxa from the seine data
seine_taxa <- kb_seine %>%
  dplyr::select(ScientificName) %>%
  distinct() %>%
  pull(ScientificName)

# Extract unique taxa from the eDNA data for April and September with Count > 0
edna_taxa <- kb_ASV_table %>%
  filter((grepl("April", SampleID) | grepl("September", SampleID)) & Count > 0) %>%
  dplyr::select(sum_taxa.x) %>%
  distinct() %>%
  pull(sum_taxa.x)

# Create a list of the taxa
taxa_list <- list(Seine = seine_taxa, eDNA = edna_taxa)

# Create the Venn diagram with labels
venn_plot <- ggVennDiagram(taxa_list, label_alpha = 0) +
  ggtitle("Venn Diagram of Unique Taxa in Seine and eDNA Samples") +
  theme(legend.position = "none") # Remove the legend for clarity
ggplot2::ggsave("/Users/marisgoodwin/Documents/GITHUB/AMSS_2025/outputs/seine_edna_venn_diagram.pdf")
# Extract the unique and shared taxa
unique_seine <- setdiff(seine_taxa, edna_taxa)
unique_edna <- setdiff(edna_taxa, seine_taxa)
shared_taxa <- intersect(seine_taxa, edna_taxa)

```

# Rarefaction curves for the eDNA data. Was our sampling depth enough to cover all the diversity within our samples 
```{r}
# this is the line of code we will be using to plot rarefaction curves for  the data. From this link: https://grunwaldlab.github.io/metacoder_documentation/workshop--06--quality_control.html. We will need to make some modifications to fit with our given data. and we want to 
rarecurve(t(obj$data$otu_counts[, "M1981P563"]), step = 20,
          sample = min(colSums(obj$data$otu_counts[, sample_data$SampleID])),
          col = "blue", cex = 1.5)
```



# Select the relevant columns
df <- edna_meta %>% dplyr::select(Region, Site, SamplingDate)

# Drop duplicate rows
df <- df %>% distinct()

# Sort the dataframe by Region and Site
df <- df %>% arrange(Region, Site)